\documentclass[a4paper,landscape,10pt]{article}
\usepackage[margin=2cm]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[lowtilde]{url}
\usepackage{color}
\usepackage[colorlinks]{hyperref}
\usepackage{scalerel}
\usepackage[numbers,sort&compress]{natbib}
\bibliographystyle{plainnat}
\usepackage[swedish,finnish,french,spanish,english]{babel}

\newcommand{\tblock}[5]{\noindent\begin{minipage}[t]{0.18\textwidth}\foreignlanguage{english}{#1}\end{minipage}\hskip 0.025\textwidth\begin{minipage}[t]{0.18\textwidth}\foreignlanguage{swedish}{#2}\end{minipage}\hskip 0.025\textwidth\begin{minipage}[t]{0.18\textwidth}\foreignlanguage{finnish}{#3}\end{minipage}\hskip 0.025\textwidth\begin{minipage}[t]{0.18\textwidth}\foreignlanguage{french}{#4}\end{minipage}\hskip 0.025\textwidth\begin{minipage}[t]{0.18\textwidth}\foreignlanguage{spanish}{#5}\end{minipage}}

%\addtolength{\oddsidemargin}{-1in}
%\addtolength{\evensidemargin}{-1in}
%\addtolength{\textwidth}{2in}



%opening
\title{{\huge \textbf{U.~S.\ Bill of Rights}}\\
	in the original English with translations\\
	to Swedish, Finnish, and French}
\author{Justina Colmena\footnote{\url{https://colmena.biz/~justina/}, \url{mailto:justina@colmena.biz}}}

\begin{document}
	
\sloppy
\maketitle
\begin{abstract}
	{\color{red}\textbf{WORK IN PROGRESS.}} An official transcript of the Bill of Rights is at the National Archives \cite{national-archives}.
	
	Note on translation:  I have found that the most critical operative word, and perhaps the most difficult to translate, is ``shall.''
	
	In Swedish this is the auxiliary verb \foreignlanguage{swedish}{<<~skola,~>>} a rather antiquated infinitive form, totally lacking in English.  The verb \foreignlanguage{swedish}{<<~skola~>>} is conjugated \foreignlanguage{swedish}{<<~skall~>>} in the present tense of the indicative mood when its subject is singular in number, and \foreignlanguage{swedish}{<<~skola~>>} for plural subjects, except \foreignlanguage{swedish}{<<~I skolen~>> (=~<<~ni ska.~>>)}  The (present indicative) first or third person plural conjugation of \foreignlanguage{swedish}{<<~skola~>>} is also spelled \foreignlanguage{swedish}{<<~skola,~>>} but it is pronounced with a falling tone on the final syllable, whereas the infinitive form is pronounced with a rising tone.  Colloquially, the infinitive form has all but disappeared, and \foreignlanguage{swedish}{<<~ska~>>} is used as the present indicative form for all persons and numbers to express the future tenses of the main verb.  I mention this because \foreignlanguage{swedish}{<<~skall,~>>} while it may sound more formal, is never correct for a plural subject in Swedish.
		
	In Finnish the auxiliary verb ``shall'' is translated \foreignlanguage{finnish}{<<~pitää.~>>}  Somewhat as in English, \foreignlanguage{finnish}{<<~pitää~>>} as an auxiliary verb is not conjugated for person and number.  Unlike English, its subject is placed in the genitive case, and in this sense \foreignlanguage{finnish}{<<~pitää~>>}  always takes the regular ending of the third person singular for any mood and tense.  If \foreignlanguage{finnish}{<<~pitää~>>} is used to express an injunction or obligation or definite future, (something that absolutely ``must,'' \foreignlanguage{finnish}{<<~täytyä,~>>} take place,) then the main verb is placed in the instructive case of the third infinitive, with the ending \foreignlanguage{finnish}{<<~\mbox{-män/-man,}~>>} N.B.\ not the illative case \foreignlanguage{finnish}{<<~\mbox{-mään/-maan,}~>>} which would be a ghastly grammatical error in this sense.

	French has a future tense integral with the conjugation of the verb. We cannot do better to translate the word ``jeopardy'' than to simply sound out the word in French: \foreignlanguage{french}{<<\,jeu-parti,\,>>} that is, a game (\foreignlanguage{french}{<<\,jeu\,>>}) of chance, where something of great value is at stake \foreignlanguage{french}{<<\,[à être] parti,\,>>} that is, ``[to be] parted,'' as the fool and his money of the common proverb;\footnote{--- ``A fool and his money are soon parted.''} and here we speak of life and limb.  Cf{.} \cite{wp-bor-fr,jpfo-bor-fr}.
	
	A fourth translation, to the Spanish language, is contemplated, and would be desirable, as it would have great relevance within the United States for a large number of Spanish-speaking citizens, but this translator is not at this time sufficiently skilled in that language to attempt it right-off. In the meantime Spanish-speaking readers are referred, rather cynically, to others' work \cite{jpfo-bor-es,wp-bor-es} in that regard. We cannot say that we are impressed with the title and authorship of that particular work, let alone the quality of the work itself, which we are not in a position to judge.
\end{abstract}

\section*{Copyright}

\begin{quote}
The work of this translation is original and is hereby placed in the public domain by the translator. Source \cite{src} is available as well as a compiled PDF copy of the work \cite{doc}. Matters of attribution for this work are academic, and are not intended to be enforced by copyright law. This translator has in the past attempted to contribute some of this work to \cite{wp-bor-sv} and \cite{wp-bor-fi}. Readers are referred to \cite{jpfo-bor} for related work, but this translator is not affiliated with that group, nor is this work based on that work.

	The selection of words, phrases, and alternative synonyms is arduous, and based on the consultation of many different dictionaries and translation services available especially online, and examples of the use of such words and phrases especially in various native-language domestic as well as diplomatic legal contexts. Older and better established alternative phraseology is generally preferred in keeping with the thinking and intentions of the Framers at the time period of the original, even if it would tend to sound somewhat archaic in some cases these days.

	We are well aware that native speakers of the modern twitter generation will snicker and giggle at words such as \foreignlanguage{swedish}{<<\,samlag\,>>} in Swedish, \foreignlanguage{finnish}{<<\,olkoon\,>>} in Finnish, and \foreignlanguage{french}{<<\,fussent\,>>} in French, but we refuse to spell ``Congress'' with a `K' or make other concessions to city slang or sloppy speech which yield to a sloppy administration of justice.
\end{quote}

\section*{Preamble}


\tblock
{Congress of the United States, begun and held at the City of New-York, on Wednesday the fourth of March, one thousand seven hundred and eighty nine.}
{Samlag av de Förenta Staterna, börjat och hållet å Staden New-York, på onsdag den fjärde mars, ett tusen sju hundra och åttio nio.}
{Eduskunta Yhdys-Valloista, alkanut ja pidetty Kaupungissa New-York keski-viikkona neljäntenä maaliskuuta, yksi tuhat seitsemän sataa ja kahdeksan kymmentä yhdeksän.}
{Le Congrès des États-Unis, commencé et tenu à la ville de New-York, mercredi le quatrième mars, mille sept cent quatre-vingt neuf.}
{!`no habla español!}

~

\tblock
{\textsc{The} Conventions of a number of the States, having at the time of their adopting the Constitution, expressed a desire, in order to prevent misconstruction or abuse of its powers, that further declaratory and restrictive clauses should be added:}
{\textsc{Som} Konventen från ett antal Stater har, samtidigt med att antaga Konstitutionen, uttryckt ett begär, i syfte att förhindra missbyggning eller missbruk av dess makter, att vidare förklarande och begränsande klausuler skulle tilläggas:}
{\textsc{Sopimuksien} lukumäärästä Valloista, ollessa Perustuslain hyväksymisensä aikana, ilmaistaneet halua, estääkseen sen valtuuksien väärää rakentamista eli väärää käyttöä, että vielä enempiä julistavia ja rajoittavia pykäliä lisättäisiin:}
{\textsc{Les Conventions} d'un nombre des États, ayant au temps d'adopter la Constitution, exprimé un désir, pour prévenir la construction érronée ou l'abuse de ses pouvoirs, que plus encore de clauses déclaratives et restrictives fussent ajoutées :}
{!`no habla español!}

~

\tblock
{And as extending the ground of public confidence in the Government, will best ensure the beneficient ends of its institution.}
{Och som utvidgande grunden för allmän tillit i Regeringen, vill bäst försäkra dess inrättnings välgörande ändamål.}
{Ja että laajentaminen perustetta julkiselle luottamukselle Hallitukseen, parhaiten varmistaisi sen perustuksen hyvää tekevän pään.}
{Et comme étendre le fond de la confiance publique en le Gouvernement, le mieux assurerait les buts bénéficiants de son institution.}
{!`no habla español!}

~

\tblock
{\textsc{Resolved} by the Senate and House of Representatives of the United States of America,}
{\textsc{Vare sig beslutat} av Amerikas Förenta Staters Senat och Representanthus,}
{\textsc{Olkoon päätetty} Ylähuoneelta ja Edustajainhuoneelta Amerikan Yhdys-Valloista,}
{\textsc{Qu'il soit résolu} au Senate et à la Chambre des Représentatives des États-Unis de l'Amérique,}
{!`no habla español!}

~

\tblock
{in Congress assembled,}
{i Samlag församlade,}
{Eduskunnassa kokoonnuttu,}
{en Congrès assemblés,}
{!`no habla español!}

~

\tblock
{two thirds of both Houses concurring,}
{två tredjedelar båda Husen konkurrerande,}
{kahden kolmannen osan molemmista Huoneista ollessa samaa mieltää,}
{deux troisièmes de toutes deux Chambres convenant,}
{!`no habla español!}

~

\tblock
{that the following Articles be proposed to the Legislatures of the several States, as amendments to the Constitution of the United States, all, or any of which Articles, when ratified by three fourths of the said Legislatures, to be valid to all intents and purposes, as part of the said Constitution; viz.}
{att de följande Artiklarna föreslås till de skilda Staternas Lagstiftande församlingar, som tillägg till de Förenta Staternas Konstitution, alla, eller några av vilka artiklar, när ratificerade av tre fjärdedelar nämnda Lagstiftande församlingar, till att gälla i alla uppsåt och avsikter som del av nämnda Konstitution; nämligen}
{että seuraavat Artiklat ehdotettaisiin kunkin Vallan lainsäädäntäelimelle, lisäyksiksi Yhdys-Valtojen Perustuslakiin, kaikkien, tai minkään joista Artikloista, ollen hyväksytty kolmelta neljänneltä osalta mainituista lainsäädäntäelimistä, kelvata kaikkiin tarkoituksiin, osana mainittua Perustuslakia; nimittäin}
{que les Articles suivants fussent proposés aux Législatures des États distincts, comme amendements à la Constitution des États-Unis, tous, ou quelques desquels articles, lorsque ratifiés par trois quatrièmes des Législatures mentionnées, pour être valides à toutes fins et intentions comme partie de la Constitution mentionnée; à savoir}
{!`no habla español!}

~

\tblock
{\textsc{Articles} in addition to, and Amendment of the Constitution of the United States of America, proposed by Congress, and ratified by the Legislatures of the several States, pursuant to the fifth Article of the original Constitution.}
{\textsc{Artiklar} i tillägg till, och Förändring av Amerikas Förenta Staters Konstitution, föreslagna av Samlaget, och ratificerade av de skilda Staternas Lagstiftande församlingar, enligt den ursprungliga Konstitutionens femte Artikel.}
{\textsc{Artiklat} lisäykseksi ja muuttamiseksi Amerikan Yhdys-Valtojen Perustuslakiin, ehdotetut Eduskunnalta, ja hyväksyttävät kunkin Vallan lainsäädäntäelimeltä, alkuperäisen Perustuslain viidennen Artiklan mukaan.}
{\textsc{Articles} en ajout et en Amendement à la Constitution des États-Unis, proposés par le Congrès, et ratifiés par les Léglislatures des États distincts, en application du cinquième Article de la Constitution originelle.}
{!`no habla español!}

\section*{Bill of Rights}
\begin{quote}\small\selectlanguage{english}
Note: The following text is a transcription of the first ten amendments to the Constitution in their original form. These amendments were ratified December 15, 1791, and form what is known as the ``Bill of Rights.''
\end{quote}
\subsection*{Amendment I}
\tblock
{Congress shall make no law}
{Samlaget skall icke stifta någon lag}
{Ei Eduskunnan pidä säättämän lakia}
{Le Congrès ne fera point de loi}
{!`no habla español!}

~

\tblock
{respecting an establishment of religion, or prohibiting the free exercise thereof;}
{avseende ett fastställande av religion, eller förbjudande den fria utövningen därav;}
{koskien uskonnon perustamista, taikka kieltäen sen vapaata noudattamista;}
{respectant un établissement de religion, ni prohibant son exercice libre;}
{!`no habla español!}

~

\tblock
{or abridging}
{eller förkortande}
{eikä lyhentäen}
{ni abrégeant}
{!`no habla español!}

~

\tblock
{the freedom of speech, or of the press;}
{yttrandefriheten, eller pressfriheten;}
{puheen tai painon vapautta;}
{la liberté de parole, ou de la presse;}
{!`no habla español!}

~

\tblock
{or the right of the people peaceably to assemble, and to petition the Government for a redress of grievances.}
{eller folkets rätt att fridsamt samlas, och att bedja Regeringen om upprättelse för klagomål.}
{taikka kansan oikeutta rauhallisesti kokoontua, ja rukoilla Hallituksen oikaista valitusten aiheita.}
{ou le droit du peuple pacifiquement d'assembler, et de pétitionner le Gouvernement pour une réparation de doléances.}
{!`no habla español!}


\subsection*{Amendment II}

\begin{quote}\small
	Note: The Second Article of Amendment to the U.S.~Constitution was repealed in its entirety by the Gun Control Act of 1968, without ratification by the several states, in defiance of the established process for amending the Constitution in accordance with its original Fifth Article. Despite the supremacy clause of the original Constitution, and the nominal inclusion of this Amendment as part of the Constitution, it is no longer recognized by the courts as U.S.~law except as an historical artifact.
\end{quote}


\tblock
{A well regulated Militia, being necessary to the security of a free State,}
{Som en väl reglerad milis är nödvändig för en fri Stats säkerhet,}
{Hyvin säännöstetyn Miliisin ollessa tarpeellista vapaan Vallan turvallisuuteen,}
{Une Milice bien réglée, étant nécessaire à la sécurité d'un État libre,}
{!`no habla español!}

~

\tblock
{the right of the people to keep and bear Arms, shall not be infringed.}
{skall Folkets rätt att inneha och bära vapen, icke inskränkas.}
{kansan oikeutta omistaa ja kantaa Aseita, ei pidä loukattaman.}
{le droit du peuple de tenir et de porter des Armes, ne sera pas enfreint.}
{!`no habla español!}

\subsection*{Gun Control Act of 1968}

\begin{quote}
	Note: As codified \cite{gpo-18usc922-2011,gpo-18usc922-2015,cornell-18usc922-1992}.
\end{quote}

\subsubsection*{as codified 18 U.S.~Code \S 922 \P(d)(4)}

\tblock
{(d) It shall be unlawful for any person to sell or otherwise dispose of any firearm or ammunition to any person knowing or having reasonable cause to believe that such person ---}
{(d) Det skall vara olagligt av någon person att sälja eller annorlunda föryttra något eldvapen eller någon ammunition till någon person vetande eller med skälig grund att tro att sådan person ---}
{(d) ---}
{(d) ---}
{(d) !`no habla español!}

~

\tblock
{(4) has been \textbf{adjudicated as a mental defective} or has been \textbf{committed to any mental institution}; ---}
{(4) har blivit \textbf{avkunnad som en mental defekt} eller har blivit \textbf{tvångsintagen till någon mental sjukvårdsinrättning}; ---}
{(4) ---}
{(4) ---}
{(4) !`no habla español!}


\subsubsection*{as codified 18 U.S.~Code \S 922 \P(g)(4)}


\tblock
{(g) It shall be unlawful for any person ---}
{(g) Det skall vara olagligt av någon person ---}
{(g) ---}
{(g) ---}
{(g) !`no habla español!}

~

\tblock
{(4) who has been \textbf{adjudicated as a mental defective} or who has been \textbf{committed to any mental institution}; ---}
{(4) som har blivit \textbf{avkunnad som en mental defekt} eller som har blivit \textbf{tvångsintagen till någon mental sjukvårdsinrättning}; ---}
{(4) ---}
{(4) ---}
{(4) !`no habla español!}

~

\tblock
{to ship or transport in interstate or \textbf{\textit{foreign commerce}}, or possess in or affecting \textbf{\textit{commerce}}, any firearm or ammunition; or to receive any firearm or ammunition which has been shipped or transported in interstate or \textbf{\textit{foreign commerce}}.}
{att skicka eller befordra i mellanstatlig eller \textbf{\textit{främmande handel}}, eller att innehava i eller rörande \textbf{\textit{handel}}, något eldvapen eller någon ammunition; eller att mottaga något eldvapen eller någon ammunition som har skickats eller befordrats i mellanstatlig eller \textbf{\textit{främmande handel}}.}
{---}
{---}
{!`no habla español!}

\subsection*{Amendment III}
\tblock
{No Soldier shall, in time of peace be quartered in any house, without the consent of the Owner, nor in time of war, but in a manner to be prescribed by law.}
{Ingen soldat skall, i tid av fred inkvarteras i något hus, utan samtycke av \textsc{ä}garen, och icke heller i tid av krig, men på ett sätt som skall föreskrivas av lag.}
{Ei sotamiestä pidä, rauhan aikana majoitettaman huoneessa, ilman Omistajan hyväksymistä, eikä sodan aikana, mutta lailta määritettävällä tavalla.}
{Aucun soldat, en temps de paix, ne sera caserné dans aucune maison, sans le consentement du Propriétaire, ni en temps de guerre, mais dans une manière à être prescrite par loi.}
{!`no habla español!}

\subsection*{Amendment IV}
\tblock
{The right of the people to be secure in their persons, houses, papers, and effects, against unreasonable searches and seizures, shall not be violated,}
{Folkets rätt att vara säkert i sina personer, hus, papper, och effekter, mot oskäliga rannsakningar och gripanden, skall icke överträdas,}
{Kansan oikeutta pysyä turvallisena henkilöissään, huoneissaan, papereissaan, ja tavaroissaan, kohtuuttomia tutkintoja ja takavarikoita vastaan, ei pidä rikottaman,}
{Le droit du peuple d'être sûr dans ses personnes, dans ses maisons, dans ses papiers, et dans ses effets, contre les cherches et les saisies non raisonnables, ne sera point violé,}
{!`no habla español!}

~

\tblock
{and no Warrants shall issue, but upon probable cause, supported by Oath or affirmation, and particularly describing the place to be searched, and the persons or things to be seized.}
{och inga Bemyndiganden skola utgå, utom på skälig grund, stödda av Ed eller försäkran, och noggrant beskrivande den plats som skall rannsakas, och de personer eller ting som skola gripas.}
{eikä Määräyksiä pidä käymän ulos, vaan todennäköisestä syystä, Valalla eli vakuutuksella tuettuja, ja erityisesti kuvaten tutkittavan paikan, ja takavarikoitavien henkilöiden tai tavaroiden.}
{et nuls Mandats ne seront émis, que de cause probable, supportés de serment ou d'affirmation, et particulièrement décrivant le lieu à être cherché et les personnes ou les choses à être saisies.}
{!`no habla español!}

\subsection*{Amendment V}
\tblock
{No person shall be held to answer for a capital, or otherwise infamous crime, unless on a presentment or indictment of a Grand Jury,}
{Ingen person skall hållas till svars för ett med dödsstraff belagt, eller på annat sätt infamt brott, såvida icke på ett föreläggande eller åtal av en Åtalsjury,}
{Ei henkilöä pidä pidettämän vastuussa kuolemalla rangaistettavasta taikka toisaalta törkeästä rikoksesta, ellei Suurelta Valamiehistöltä syytä löydettämisellä eli syytteeseen panettamisella,}
{Nulle personne ne sera mise à la question, d'un crime capital ou d'ailleurs infâme, sinon à la présentation ou à la mise en cause d'un Grand Jury,}
{!`no habla español!}

~

\tblock
{except in cases arising in the land or naval forces, or in the Militia, when in actual service in time of War or public danger;}
{utom i fall uppstående inom land- eller sjöstridskrafterna, eller inom Milisen, när i pågående tjänst i tid av krig eller allmän fara;}
{sitä paitsi tapauksissa noustaen ylös maan tai meren voimissa, tai Miliisissä, ollen nykyisessä palveluksessa Sodan tai julkisen vaaran aikana;}
{excepté en cas survenant dans les forces armées ou navales, ou dans la Milice, quant à service actuelle en temps de guerre ou de danger public;}
{!`no habla español!}

~

\tblock
{nor shall any person be subject for the same offence to be twice placed in jeopardy of life or limb;}
{icke heller skall någon person stå för samma brott till att två gånger utsättas för fara för liv eller lem;}
{eikä pidä mitään henkilöä altistettaman samasta rikoksesta kahdesti panettaa vaaraan elämästä taikka jäsenestä;}
{ni ne sera assujettie nulle personne de la même offense à être deux fois mise au <<\,jeu-parti\,>> de la vie ou du membre;}
{!`no habla español!}

~

\tblock
{nor shall be compelled in any criminal case to be a witness against himself,}
{icke heller skall tvingas i något brottmål att vara vittne mot sig själv,}
{eikä pidä pakotettaman mistäkään rikossyytteestä olemaan vierasmies itseänsä vastaan;}
{ni sera contrainte en aucun cas criminel à être témoin contre elle-même;}
{!`no habla español!}

~

\tblock
{nor be deprived of life, liberty, or property, without due process of law;}
{och icke heller berövas liv, frihet, eller egendom, utan vederbörlig rättegång;}
{eikä häneltä riistettämän elämää, vapautta, taikka omaisuutta, ilman asianmukaista oikeudenkäyntiä;}
{ni privée de la vie, de la liberté, ou de la propriété, sans le procès dû de la loi;}
{!`no habla español!}

~

\tblock
{nor shall private property be taken for public use, without just compensation.}
{icke heller heller skall privat egendom tagas för det allmänna nyttjandet, utan rättvis kompensation.}
{eikä pidä yksityistä omaisuutta otettaman julkiseen käyttöön, ilman oikeudenmukaista korvausta.}
{ni ne sera prise la propriété privée pour l'usage public, sans juste compensation.}
{!`no habla español!}


\subsection*{Amendment VI}
\tblock
{In all criminal prosecutions, the accused shall enjoy the right}
{I alla brottmål, skall den anklagade njuta rätten}
{Kaikissa rikossyytteissä, pitää syytetyn nauttiman oikeutta}
{En toutes accusations criminelles, l'accusé jouira du droit}
{!`no habla español!}

~

\tblock
{to a speedy and public trial, by an impartial jury of the State and district wherein the crime shall have been committed, which district shall have been previously ascertained by law,}
{till en snabb och publik prövning, av en opartisk jury av Staten och distriktet däri brottet skall hava begåtts, vilket distrikt skall hava redan förut bestämts av lag,}
{vauhdikkaaseen ja julkiseen koetukseen, tasapuoliselta valamiehistöltä siitä Vallasta ja piiristä, joissa rikokseen olisi syyllistytty, mikä piiri olisi jo ennestään lailta määrätty,}
{d'une épreuve prompte et publique, par un jury impartial de l'État et du département dans lesquels le crime aurait été commis, lequel département aurait été déjà précédemment vérifié par la loi,}
{!`no habla español!}

~

\tblock
{and to be informed of the nature and cause of the accusation;}
{och att bliva informerad om anklagelsens natur och grund;}
{ja tiedostaa syytteen luontoa ja syytä;}
{et d'être informé de la nature et de la cause de l'accusation;}
{!`no habla español!}

~

\tblock
{to be confronted with the witnesses against him;}
{att konfronteras med vittnena emot honom;}	
{vastustaa vieraitamiehiä häntä vastaan;}
{d'être confronté avec les témoins contre lui;}
{!`no habla español!}

~

\tblock
{to have compulsory process for obtaining witnesses in his favor,}
{att begagna sig av tvingande process för att skaffa vittnen i sin favör,}
{Hyödyntyä pakollista käyntiä hankkia vieraitamiehiä hänen tähtensä,}
{d'avoir du procès contraignant pour obtenir des témoins à sa faveur,}
{!`no habla español!}

~

\tblock
{and to have the Assistance of Counsel for his defence.}
{och att hava Medverkan av Rättegångsbiträde till sitt försvar.}
{ja Neuvon Apua hänen puolustukselleen.}
{et d'avoir l'Assistance du Conseil pour sa défense.}
{!`no habla español!}


\subsection*{Amendment VII}
\tblock
{In Suits at common law, where the value in controversy shall exceed twenty dollars, the right of trial by jury shall be preserved,}
{I civilrättsmål, där värdet i tvist skall överstiga tjugo dollar, skall rätten till prövning av jury bevaras,}
{Asioissa tapaoikeudella, joissa arvo riitasta ylittäisi kahtakymmentä dollaria, oikeutta koetukseen valamiehistöltä pitää varjeltaman,}
{En actions en justice commune, où la valeur en controverse excéderait vingt dollars, le droit de l'épreuve par jury sera préservé,}
{!`no habla español!}

~

\tblock
{and no fact tried by a jury, shall be otherwise re-examined in any Court of the United States, than according to the rules of the common law.}
{och intet faktum prövat av en jury, skall annorlunda åter sättas i fråga inför någon domstol inom de Förenta Staterna, än enligt civilrättens förfaranden.}
{eikä tosiasiaa, koetettua valamiehistöltä, pidä toisaalta uudelleen tutkettaman minkään Tuomioistuimen edessä Yhdys-Valloista, kuin tapaoikeuden ettelyjen mukaisesti.}
{et nul fait, éprouvé par un jury, ne sera d'ailleurs réexaminé devant aucune cour de l'États-Unis, que d'après les règles de la justice commune.}
{!`no habla español!}


\subsection*{Amendment VIII}
\tblock
{Excessive bail shall not be required,}
{Omåttlig borgensumma skall icke krävas,}
{Liiallista takausta ei pidä vaadittaman,}
{Nulle caution excessive ne sera point requérie,}
{!`no habla español!}

~

\tblock
{nor excessive fines imposed,}
{icke heller omåttliga böter åläggas,}
{eikä liiallisia sakkoja määrättämän,}
{ni des amendes excessives imposées,}
{!`no habla español!}

~

\tblock
{nor cruel and unusual punishments inflicted.}
{och icke heller grymma och ovanliga straff åsamkas.}
{eikä julmia ja epätavallisia rangaistuksia langetettaman.}
{ni des punitions cruelles et anormales infligées.}
{!`no habla español!}

\subsection*{Amendment IX}
\tblock
{The enumeration in the Constitution, of certain rights, shall not be construed to deny or disparage others retained by the people.}
{Uppräkningen i Konstitutionen, av vissa rättigheter, skall icke tolkas till att förneka eller ringakta andra behållna av folket.}
{Laskemisesta Perustuslaissa, eräitä oikeuksia, ei pidä rakennettaman kieltää eli väheksyä toisia, kansalle säilytettyjä.}
{L'énumération dans la Constitution, de certains droits, ne sera construite à nier ni à décrier des autres retenus par le peuple.}
{!`no habla español!}

\subsection*{Amendment X}
\tblock
{The powers not delegated to the United States by the Constitution, nor prohibited by it to the States, are reserved to the States respectively, or to the people.}
{De makter, icke delegerade till de Förenta Staterna av Konstitutionen, och icke heller förbjudna av den till Staterna, reserveras till Staterna respektive, eller till folket.}
{Valtuudet, jotka ei ole siirretty Yhdys-Valloille Perustuslailta, eikä joista siltä ole kielletty Valtoja, varataan kullekin Vallalle, tai kansalle.}
{Les pouvoirs non délégués aux États-Unis par la Constitution, ni prohibés par elle aux États, sont réservés aux États distincts, ou au peuple.}
{!`no habla español!}


\section*{Civil Rights}
\subsection*{Amendment XIII}
\begin{quote}\small
	Passed by Congress January 31, 1865. Ratified December 6, 1865.
\end{quote}
\subsubsection*{Section 1.}
\tblock
{Neither slavery nor involuntary servitude,}
{Varken träldom eller ofrivillig betjäntskap,}
{Ei orjuuden, eikä pakollisen palveluksen,}
{Ni l'esclavage ni la servitude involontaire,}
{!`no habla español!}

~

\tblock
{except as a punishment for crime whereof the party shall have been duly convicted,}
{såvida icke som straff för brott därav parten skall hava vederbörligen fällts,}
{sitä paitsi rangaistukseksi rikoksesta, johonka osa olisi asianmukaisesti syylliseksi todettu,}
{excepté comme punition d'un crime, duquelle la partie serait dûment convaincue,}
{!`no habla español!}

~

\tblock
{shall exist within the United States, or any place subject to their jurisdiction.}
{skall bliva till inom de Förenta Staterna eller någorstädes inom deras domsrätt.}
{pidä tuleman olemaan Yhdys-Valtojen sisälle, taikka mihinkään paikkaan niiden toimivallalla.}
{n'existera au-dedans des États-Unis, ni à nulle part assujettie à leur juridiction.}
{!`no habla español!}


\subsubsection*{Section 2.}
\tblock
{Congress shall have power to enforce this article by appropriate legislation.}
{Samlaget skall hava makt att framtvinga denna artikel genom tillämplig lagstiftning.}
{Eduskunnalla pitää oleman valtuuden pakottaa tätä artiklaa asiallisen lainsäädönnön kautta.}
{Le Congrès aura le pouvoir d'appliquer cet article-ci par législation convenable.}
{!`no habla español!}



\subsection*{Amendment XIV}
\begin{quote}\small
	Passed by Congress June 13, 1866. Ratified July 9, 1868.
\end{quote}
\subsubsection*{Section 1.}
\tblock
{All persons born or naturalized in the United States, and subject to the jurisdiction thereof, are citizens of the United States and of the State wherein they reside.}
{Alla personer födda eller naturaliserade i de Förenta Staterna, och utsatta för rättskipningen darav, är medborgare av de Förenta Staterna och av staten däri de bor.}
{Kaikki henkilöt, syntyneet tai kansalaistetut Yhdys-Valtoihin, ja alistetut niiden oikeudenkäyttöön, ovat kansalaiset Yhdys-Valtoista ja siitä Vallasta, jossa he asuvat.}
{Toutes personnes nées ou naturalisées aux États-Unis, et assujetties à leur juridiction, sont citoyens des États-Unis et de l'État dans lequel elles résident.}
{!`no habla español!}

~

\tblock
{No State shall make or enforce any law which shall abridge the privileges or immunities of citizens of the United States;}
{Ingen stat skall stifta eller påtvinga någon lag som skall förkorta de privilegier eller immuniteter tillhörande medborgarna av de Förenta Staterna;}
{Ei Vallan pidä säättämän eikä pakottaman lakia, jonka pitisi lyhentämän Yhdys-Valtojen kansalaisten etuoikeuksia eli koskemattomuuksia;}
{Nul État ne fera ni appliqera aucune loi, qui abrégerait des privilèges ou des immunités des citoyens des États-Unis;}
{!`no habla español!}

~

\tblock
{nor shall any State deprive any person of life, liberty, or property, without due process of law;}
{icke heller skall någon stat beröva någon person liv, frihet, eller egendom, utan vederbörlig rättegång;}
{eikä pidä minkään Vallan riistämän miltäkään henkilöltä elämää, vapautta, taikka omaisuutta, ilman asianmukaista oikeudenkäyntiä;}
{ni aucune personne ne privera aucun État de la vie, de la liberté, ou de la propriéte, sans le procès dû de la loi;}
{!`no habla español!}

	
~

\tblock
{nor deny to any person within its jurisdiction the equal protection of the laws.}
{och icke heller förneka någon person inom sin domsrätt lagens lika skydd.}
{eikä kieltämän miltäkään henkilöltä sen toimivallalla lain samaa suojaa.}
{ni nier à nulle personne sous sa juridiction la protection égale des lois.}
{!`no habla español!}

~
		
$\cdots$

\subsubsection*{Section 5.}
\tblock
{The Congress shall have the power to enforce, by appropriate legislation, the provisions of this article.}
{Samlaget skall hava makten att framtvinga, genom tillämplig lagstiftning, de åtgärder utsatta i denna artikel.}
{Eduskunnalla pitää oleman valtuuden pakottaa, asiallisen lainsäädönnön kautta, tämän artiklan toimenpiteitä.}
{Le Congrès aura le pouvoir, par législation convenable, d'appliquer les dispositions de cet article-ci.}
{!`no habla español!}


\subsection*{Amendment XV}
\begin{quote}\small
	Passed by Congress February 26, 1869. Ratified February 3, 1870.
\end{quote}

\subsubsection*{Section 1.}
\tblock
{The right of citizens of the United States to vote shall not be denied or abridged by the United States or by any State}
{De Forenta Staternas medborgares rätt att rösta skall icke förnekas eller förkortas av de Förenta Staterna eller av någon Stat}
{Yhdys-Valtojen kansalaisten oikeutta äänestää ei pidä kiellettämän eikä lyhennettämän Yhdys-Valloilta taikka miltäkään vallalta}
{Le droit des citoyens des États-Unis ne sera nié ni abrégé par les États-Unis ni par aucun État}
{!`no habla español!}

~

\tblock
{on account of race, color, or previous condition of servitude~---}
{på grund av ras, färg, eller förutvarande tillstånd av tjänedom~---}
{selityksellä rodusta, väristä, taikka edelliseltä palveluksen ehdolta~---}
{à compte de race, de couleur, ou de condition de servitude précédente~---}
{!`no habla español!}

\subsubsection*{Section 2.}
\tblock
{The Congress shall have the power to enforce this article by appropriate legislation.}
{Samlaget skall hava makten att framtvinga denna artikel genom tillämplig lagstiftning.}
{Eduskunnalla pitää oleman valtuuden pakottaa tätä artiklaa asiallisen lainsäädönnön kautta.}
{Le Congrès aura le pouvoir d'appliquer cet article-ci par législation convenable.}
{!`no habla español!}

\subsection*{Amendment XIX}
\begin{quote}\small
Passed by Congress June 4, 1919. Ratified August 18, 1920.
\end{quote}
\tblock
{The right of citizens of the United States to vote shall not be denied or abridged by the United States or by any State}
{De Forenta Staternas medborgares rätt att rösta skall icke förnekas eller förkortas av de Förenta Staterna eller av någon Stat}
{Yhdys-Valtojen kansalaisten oikeutta äänestää ei pidä kiellettämän eikä lyhennettämän Yhdys-Valloilta taikka miltäkään vallalta}
{Le droit des citoyens des États-Unis ne sera nié ni abrégé par les États-Unis ni par aucun État}
{!`no habla español!}

~

\tblock
{on account of sex.}
{på grund av kön.}
{selityksellä sukupuolesta.}
{à compte de sexe.}
{!`no habla español!}

~

\tblock
{Congress shall have power to enforce this article by appropriate legislation.}
{Samlaget skall hava makt att framtvinga denna artikel genom tillämplig lagstiftning.}
{Eduskunnalla pitää oleman valtuuden pakottaa tätä artiklaa asiallisen lainsäädönnön kautta.}
{Le Congrès aura du pouvoir d'appliquer cet article-ci par législation convenable.}
{!`no habla español!}


\subsection*{Amendment XXIV}
\begin{quote}\small
Passed by Congress August 27, 1962. Ratified January 23, 1964.
\end{quote}
\subsubsection*{Section 1.}
\tblock
{The right of citizens of the United States to vote in any primary or other election}
{De Förenta Staternas medbörgares rätt att rösta i något primärt eller annat val}
{Yhdys-Valtojen kansalaisten oikeutta äänestää missäkään esivaalissa taikka toisessa vaalissa}
{Le droit des citoyens des États-Unis de voter dans aucune élection primaire ou autre}
{!`no habla español!}

~

\tblock
{for President or Vice President,}
{för President eller Vice-President,}
{presidentistä taikka varapresidentistä,}
{de Président ou de Vice-Président,}
{!`no habla español!}

~

\tblock
{for electors for President or Vice President,}
{för elektorer för President eller Vice-President,}
{valitsijamiehistä presidentistä taikka varapresidentistä,}
{d'électeurs de Président ou de Vice-Président,}
{!`no habla español!}

~

\tblock
{or for Senator or Representative in Congress,}
{eller för Senator eller Representat i Samlag,}
{taikka yliherrasta taikka edustajasta Eduskunnassa,}
{ou de Sénateur ou de Representative à Congrès,}
{!`no habla español!}

~

\tblock
{shall not be denied or abridged by the United States or any State}
{skall icke förnekas eller förkortas av de Förenta Staterna eller av någon Stat}
{ei pidä kiellettämän eikä lyhennettämän Yhdys-Valloilta taikka miltäkään vallalta}
{ne sera nié ni abrégé par les États-Unis ni par aucun État}
{!`no habla español!}

~

\tblock
{by reason of failure to pay any poll tax or other tax.}
{på grund av underlåtenhet att betala någon skatt per capita eller annan skatt.}
{selityksellä olemasta maksamatta henkilömääräveroa taikka toista veroa.}
{à compte de défaillance de payer aucune capitation ou autre taxe.}
{!`no habla español!}


\subsubsection*{Section 2.}
\tblock
{The Congress shall have power to enforce this article by appropriate legislation.}
{Samlaget skall hava makt att framtvinga denna artikel genom tillämplig lagstiftning.}
{Eduskunnalla pitää oleman valtuuden pakottaa tätä artiklaa asiallisen lainsäädönnön kautta.}
{Le Congrès aura du pouvoir d'appliquer cet article-ci par législation convenable.}
{!`no habla español!}

\subsection*{Amendment XXVI}
\tblock
{The right of citizens of the United States,}
{De Forenta Staternas medborgares rätt,}
{Yhdys-Valtojen kansalaisten oikeutta,}
{Le droit des citoyens des États-Unis,}
{!`no habla español!}


~

\tblock
{who are eighteen years of age or older,}
{som äro av aderton års ålder eller äldre,}
{jotka ovat kahdeksantoistakymmentä vuotta vanhoja taikka vanhempia,}
{qui sont de dix-huit ans d'âge ou plus vieux,}
{!`no habla español!}

~

\tblock
{to vote}
{att rösta}
{äänestää}
{de voter}
{!`no habla español!}

~

\tblock
{shall not be denied or abridged by the United States or by any State on account of age.}
{skall icke förnekas eller förkortas av de Förenta Staterna eller av någon Stat på grund av ålder.}
{ei pidä kiellettämän eikä lyhennettämän Yhdys-Valloilta taikka miltäkään vallalta selityksellä vanhuudesta.}
{ne sera nié ni abrégé par les États-Unis ni par aucun État à compte d'âge.}
{!`no habla español!}

\subsubsection*{Section 2.}
\tblock
{The Congress shall have power to enforce this article by appropriate legislation.}
{Samlaget skall hava makt att framtvinga denna artikel genom tillämplig lagstiftning.}
{Eduskunnalla pitää oleman valtuuden pakottaa tätä artiklaa asiallisen lainsäädönnön kautta.}
{Le Congrès aura du pouvoir d'appliquer cet article-ci par législation convenable.}
{!`no habla español!}


\section*{The End of Civil Rights}
\subsection*{Not just psychiatry --- all sorts of medical quackery}

\tblock
{All male infants, born in the United States, must be circumcised immediately at birth.}  
{Alla barn av mankön, födda i de Förenta Staterna, måste omskaras strax på födelsen.}
{Kaikkia miespuolisia lapsia, Yhdys-Valloissa syntyneitä, täytyy ympärileikata heti syntymällä.}
{Tous enfants mâles, nés aux États-Unis, doivent être circoncis immédiatement à la naissance.}
{!`no habla español!}

\bibliography{bor}{}

\end{document}
